---
title: "How to choose a license for an academic project"
subtitle: ""
author: "Hugo Gruson"
institute: ""
date: ""
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  out.width = "80%",
  fig.width=5,
  fig.height=3.5,
  dev='svglite'
)
```

```{css}
li, p {
  font-size: 1.5rem;
  line-height: 150%;
}
.footnote {
  font-size: 1rem;
}
```

```{r}
library(fontawesome)
```

## Free/Libre Open Source Software ("FLOSS")

Four freedoms: 
- The freedom to run the program as you wish, for any purpose.
- The freedom to study how the program works, and change it so it does your computing as you wish.
- The freedom to redistribute copies so you can help others.
- The freedom to distribute copies of your modified versions to others. By doing this you can give the whole community a chance to benefit from your changes.

.footnote[Source: <https://www.gnu.org/philosophy/free-sw.en.html>]

---

## Copyleft licenses

You have to keep the same license (or a compatible FLOSS license) for downstream
projects.

- GNU Public License (version 2 & 3) and derivatives:
  * AGPL (for applications on servers mainly)
  * LGPL (for libraries mainly)
--

- Creative Commons ShareAlike ("CC-SA")

---

## Copyleft licenses

### GNU Public License notable examples

```{r}
fa("fab fa-linux", height = "5em", margin_right = "2em")
fa("fab fa-git", height = "5em", margin_right = "2em")
fa("fab fa-r-project", height = "5em")
```


--

### Creative Commons ShareAlike notable examples

```{r, class.source = "logos"}
fa("fab fa-wikipedia-w", height = "5em", margin_right = "2em")
fa("fab fa-stack-overflow", height = "5em")
```

---

## Non-copyleft licenses

- MIT
- Apache
- BSD
- CC0 / zero-clause BSD / public domain

---

## Non-copyleft licenses

### MIT notable examples

```{r, class.source = "logos"}
fa("fab fa-react", height = "3em", margin_right = "2em")
fa("fab fa-angular", height = "3em", margin_right = "2em")
fa("fab fa-rust", height = "3em", margin_right = "2em")
```

### BSD(-like) notable examples

```{r, class.source = "logos"}
fa("fab fa-python", height = "3em", margin_right = "2em")
```

numpy, pandas, scipy, pytorch

### Apache notable examples

```{r, class.source = "logos"}
fa("fab fa-android", height = "3em")
```

---

## Other notable cases

Intermediary copyleft:

- MPL: only redistribute parts under copyleft, not the whole codebase.

--

Non-free licenses:

- Creative Commons NoDerivative / NoCommercial (CC-ND / CC-NC)
--

- JSON license

> The Software shall be used for Good, not Evil.

---

## A resource to help you choose

```{r}
knitr::include_app("https://choosealicense.com")
```

.footnote[<https://choosealicense.com/>]

---

## Licenses in the R package ecosystem

As of `r Sys.Date()`:

```{r, cache = TRUE, cache.extra = Sys.Date()}
cran_today <- available.packages(contriburl = contrib.url("https://cloud.r-project.org"))
```

```{r, dev='svglite', plot = TRUE}
library(dplyr)
library(ggplot2)
cran_today %>%
  as.data.frame() %>%
  mutate(simple_license = case_when(
    grepl("^AGPL", License) ~ "AGPL",
    grepl("^GNU Affero", License) ~ "AGPL",
    grepl("^GPL", License) ~ "GPL",
    grepl("^GNU General Public License", License) ~ "GPL",
    grepl("^LGPL", License) ~ "LGPL",
    grepl("^Lesser G[nN][uU]", License) ~ "LGPL",
    grepl("GNU Lesser", License) ~ "LGPL",
    grepl("^MIT", License) ~ "MIT",
    grepl("^MPL", License) ~ "MPL",
    grepl("^Mozilla", License) ~ "MPL",
    grepl("^Apache License", License) ~ "Apache",
    grepl("^BSD(_| )2", License) ~ "BSD 2-clause",
    grepl("^BSD(_| )3", License) ~ "BSD 3-clause",
    grepl("^Artistic", License) ~ "Artistic",
    grepl("^CC0", License) ~ "Public domain",
    grepl("Unlimited", License) ~ "Public domain",
    grepl("^CC", License) ~ "Creative Commons",
    grepl("^Creative Commons", License) ~ "Creative Commons",
    TRUE ~ "Other"
  )) %>%
  count(simple_license) %>%
  ggplot(aes(y = reorder(simple_license, n), x = n)) +
    geom_histogram(stat = "identity") +
    labs(y = "Licence family", x = "Number of packages on CRAN") +
    theme_minimal()
```
---

## License of most downloaded CRAN packages

During the last month:

```{r}
rlib_pkgs <- c("vctrs", "rlang", "lifecycle", "ellipsis", "pillar", 
               "textshaping", "cli", "ragg", "glue", "cpp11", "pkgdown", 
               "crayon")
tidyverse_pkgs <- c("ggplot2", "dplyr", "tibble", "magrittr")
jeroen_pkgs <- c("jsonlite")
sf_pkgs <- c("sf")
yihui_pkgs <- c("xfun")

cranlogs::cran_top_downloads(when = "last-month", 20) %>%
  ggplot(aes(x = count, y = reorder(package, -rank))) +
  geom_histogram(stat = "identity") +
  theme_minimal() +
  labs(y = "Package", x = "Number of downloads")
```
